---
title: "Testing a ggplot theme"
description: |
  Playing with options to decide on the theme I want to use.
author:
  - name: Jo Dudding
    url: {}
date: 2020-12-13
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 19 / 2.54,
  fig.align = 'center',
  fig.asp = 0.618
  
  )
```

First we'll start with a chart using the default theme.

```{r}
library(tidyverse)
library(ggtext)

p_bar <- ggplot(mtcars, aes(drat)) +
  geom_histogram(bins = 5) +
  labs(
    x = 'Title of x',
    y = 'Title of y',
    title = 'Overall Title',
    subtitle = 'This is the subtitle',
    caption = 'May need a caption as well'
  )

p_bar
```

I tried the different base themes and this was the closest to what I wanted.

```{r}
p_bar + theme_bw()
```

Let's try changing the font

```{r, eval = FALSE}
p_bar <- p_bar + 
  theme_bw() %+replace%
  theme(
    # default font
    text = element_text(family = "Fira Sans")
  )

p_bar
```

Add some other layout changes to:

- Move the title and caption to align with the plot not the panel
- Put some extra room between the plot and the title/caption
- Make the text fields use `element_markdown`
- Right align the axis titles
- Remove the panel border
- Add in axis lines the same colour as the gridlines and change the tick marks
  to use the same colour.

```{r}
p_bar <- p_bar +
  theme(
    # align title and caption to the plot not the panel
    plot.title.position = 'plot',
    plot.caption.position = 'plot',
    # change the title and caption to markdown and move them futher from the plot
    plot.title = element_markdown(
      hjust = 0, 
      margin = margin(c(0, 0, 10, 0))
    ),
    plot.subtitle = element_markdown(
      hjust = 0, 
      margin = margin(c(0, 0, 15, 0))
    ),
    plot.caption = element_markdown(
      hjust = 1, 
      margin = margin(c(10, 0, 0, 0))
    ),
    # move axis titles to the left/top and change them to markdown
    axis.title = element_markdown(hjust = 1),
    # allow the axis values to the markdown as well
    axis.text = element_markdown(),
    # remove the panel border
    panel.border = element_blank(),
    # put in the axis lines with a slightly thicker line than the gridlines
    axis.line = element_line(colour = "grey92", size = rel(1.5)),
    # make the tickmarks the same colour
    axis.ticks = element_line(colour = "grey92")
  )

p_bar
```

Let's see what it looks like with facets.

```{r}
p_bar <- p_bar + 
  facet_wrap(~as_factor(carb))

p_bar
```

I've left aligned the text in the facet and removed the box and border.

```{r}
p_bar <- p_bar + 
  theme(
    # facet strip text left aligned with extra space above
    strip.text = element_markdown(hjust = 0, margin = margin(c(10, 0, 0, 0))),
    # clear colour and fill for strip
    strip.background = element_rect(colour = NA, fill = NA)
  )

p_bar
```

Time to add some colour.

```{r}
add_colour <- "#5F187FFF"

p_bar <- p_bar +
  theme(
    # colour the title
    plot.title = element_markdown(
      hjust = 0, margin = margin(c(0, 0, 10, 0)), colour = add_colour
    ),
    # colour the strip text
    strip.text = element_markdown(
      hjust = 0, margin = margin(c(10, 0, 0, 0)), colour = add_colour
    )
  )

p_bar
```

Nearly there.  I think I'll made the gridlines dotted.

```{r}
add_colour <- "#5F187FFF"

p_bar <- p_bar +
  theme(
    # dotted gridlines
    panel.grid = element_line(linetype = 'dotted')
  )

p_bar
```

Okay.  Let's save this as a theme with options to change out the highlight 
colour, font family and font size


```{r}
theme_jo <- function(
  add_colour = "#5F187FFF",
  base_family = "Fira Sans",
  x_grid_colour = "grey92",
  y_grid_colour = "grey92",
  ...
  ) {
  
  theme_bw(...) %+replace%
  theme(
    # default font
    text = element_text(family = base_family),
    # align title and caption to the plot not the panel
    plot.title.position = 'plot',
    plot.caption.position = 'plot',
    # change the title and caption to markdown and move them further from the plot
    plot.title = element_markdown(
      hjust = 0, 
      margin = margin(c(0, 0, 10, 0)),
      colour = add_colour
    ),
    plot.subtitle = element_markdown(
      hjust = 0, 
      margin = margin(c(0, 0, 15, 0))
    ),
    plot.caption = element_markdown(
      hjust = 1, 
      margin = margin(c(10, 0, 0, 0))
    ),
    # move axis titles to the left/top and change them to markdown
    axis.title = element_markdown(hjust = 1),
    # allow the axis values to the markdown as well
    axis.text = element_markdown(),
    # remove the panel border
    panel.border = element_blank(),
    # put in the axis lines with a slightly thicker line than the gridlines
    axis.line = element_line(colour = "grey92", size = rel(1.5)),
    # make the tickmarks the same colour
    axis.ticks = element_line(colour = "grey92"),
    # facet strip text left aligned with extra space above
    strip.text = element_markdown(
      hjust = 0, margin = margin(c(10, 0, 0, 0)), colour = add_colour
    ),
    # clear colour and fill for strip
    strip.background = element_rect(colour = NA, fill = NA),
    # dotted gridlines
    panel.grid = element_line(linetype = 'dotted'),
    # ability to use a different colour for the gridlines
    panel.grid.major.x = element_line(colour = x_grid_colour),
    panel.grid.major.y = element_line(colour = y_grid_colour)
  )
}
```

And apply it to a new chart with a markdown change in the title.

```{r, preview = TRUE}
ggplot(diamonds, aes(x=carat, y=price, color=clarity)) + 
  geom_point() +
  labs(
    title = '*Price* of **diamonds** for differing *carats*',
    subtitle = 'The *clarity* is also a driver of *price*',
    caption = 'Source: unknown'
  ) +
  theme_jo()
```

And using different parameters.

```{r}
ggplot(diamonds, aes(x=clarity, y=price)) + 
  geom_boxplot() +
  labs(
    title = '*Price* of **diamonds** for differing *carats*',
    subtitle = 'This time its a boxplot',
    caption = 'Source: unknown'
  ) +
  theme_jo(
    add_colour = '#FFA500', 
    base_family = 'sans',
    y_grid_colour = '#FFA500',
    x_grid_colour = 'white')
```