---
title: "Helper functions for factors"
description: |
  Some helper functions to let you change the character strings of a 
  factor variable without effecting the order of the levels.
author:
  - name: Jo Dudding
date: 2020-12-12
theme: theme.css
output:
  distill::distill_article:
    self_contained: false
draft: true    
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this post, I thought I'd share some of the help fuctions for factors
that I'm using. They are based on [Hadley's response to this issue for the 
stringr package](https://github.com/tidyverse/stringr/issues/107).

The helper functions are:

- `str_wrap_factor`: to add string wrapping
  ([`str_wrap`](https://stringr.tidyverse.org/reference/str_wrap.html)) 
  to a factor variable
- `str_br_factor`: similar to the above, but using \<br\> instead of \\n
- `str_replace_factor`: to use
  [`str_replace`](https://stringr.tidyverse.org/reference/str_replace.html) 
  on a factor variable
- `str_sub_factor`: to use
  [`str_sub`](https://stringr.tidyverse.org/reference/str_sub.html)
  for a factor variable

Let's start by looking to see what happens if you use these stringr functions
on a factor variable.

```{r}
library(tidyverse)

has_factor <- tibble(
    x = rep(1:10, 3),
    y = runif(30),
    fct_name = rep(c(
      'This is my first long factor', 
      'And my second long factor', 
      'But the third is the longest of all of the factors so far'
    ), 10)
  ) %>% 
  mutate(
    fct_name = factor(fct_name, levels = c(
      'This is my first long factor', 
      'And my second long factor', 
      'But the third is the longest of all of the factors so far'
    ))
  )

glimpse(has_factor)

```

You can see these are in the order they are listed in the levels statement 
above.

```{r}
fct_count(has_factor$fct_name)
```

Now, what happens when I try and apply `str_wrap`?

```{r}
fct_count(str_wrap(has_factor$fct_name, 20))
```

You can see that we've lost the ordering and its now in alphabetic order.

Here's what it looks like in a graph.

```{r}
# source the file that contains my theme first
source("../../scripts/mytheme.r")

ggplot(has_factor, 
  aes(x = str_wrap(fct_name, 20), y = y)
  ) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(x = NULL, y = 'Name of this variable') +
  theme_jo()
```

Still out of order (and also reversed as the factor order starts from zero).

The purpose of the `str_wrap_factor` helper function is to apply the wrapping
but not lose the ordering of the factor variable.

Here's the function

```{r}
str_wrap_factor <- function(x, ...) {
  levels(x) <- str_wrap(levels(x), ...)
  x
}
```

And here's what happens when we use it.

```{r}
fct_count(str_wrap_factor(has_factor$fct_name, 20))
```

The factor order has been retained.  Let's see it in use with a graph.

```{r, preview = TRUE}
ggplot(has_factor, 
  aes(x = str_wrap_factor(fct_name, 20), y = y)
  ) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(x = NULL, y = 'Name of this variable')
```

Let's try reversing the (flipped) x axis so the factor is ordered down instead 
of up.

```{r}
ggplot(has_factor, 
  aes(x = fct_rev(str_wrap_factor(fct_name, 20)), y = y)
  ) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(x = NULL, y = 'Name of this variable')
    
```

If you are using the `ggtext` package or creating a `knitr::kable` table you may find
that \\n is ignored.

```{r}
library(knitr)

has_factor %>% 
  count(fct_name, wt = y) %>% 
  mutate(fct_name = str_wrap_factor(fct_name, 20)) %>% 
  kable()
```

That doesn't look right.  Let's extend the previous function to replace \\n with
\<br\>.


```{r}
str_br_factor <- function(x, ...) {
  levels(x) <- str_wrap(levels(x), ...)
  levels(x) <- str_replace_all(levels(x), '\n', '<br>')
  x
}
```

Here's the factor count again.

```{r}
fct_count(str_br_factor(has_factor$fct_name, 20))

```

And the table using `str_br_factor`.

```{r}
has_factor %>% 
  count(fct_name, wt = y) %>% 
  mutate(fct_name = str_br_factor(fct_name, 20)) %>% 
  kable()
```

Much better.

> Still to do:

`str_replace`

`str_sub`

You may be creating grouping based on other variables and want this to be a
factor.

```{r}
calc_factor <- tibble(x = runif(30)) %>% 
  mutate(
    groups = case_when(
      x < 0.25 ~ 'First level',
      x < 0.66 ~ 'Second level',
      TRUE ~ 'Last level'
    ),
    groups = factor(groups)
  )
    
fct_count(calc_factor$groups)
```

The default when using `factor` is alpabetic order.  You could use factor and
list out the levels instead.

```{r}
calc_factor <- tibble(x = runif(30)) %>% 
  mutate(
    groups = case_when(
      x < 0.25 ~ 'First level',
      x < 0.66 ~ 'Second level',
      TRUE ~ 'Last level'
    ),
    groups = factor(groups, levels = c(
      'First level',
      'Second level',
      'Last level'
    ))
  )
  
fct_count(calc_factor$groups)
```

But this involves entering the strings twice.

```{r}
calc_factor <- tibble(x = runif(30)) %>% 
  mutate(
    groups = case_when(
      x < 0.25 ~ 'a First level',
      x < 0.66 ~ 'b Second level',
      TRUE ~ 'c Last level'
    ),
    groups = factor(groups)
  )
    
fct_count(calc_factor$groups)
```

```{r}
str_ord_factor <- function(x, start = 3L, ...) {
  x <- factor(x)
  levels(x) <- str_sub(levels(x), start = start, ...)
  x
}
```

```{r}
calc_factor <- tibble(x = runif(30)) %>% 
  mutate(
    groups = case_when(
      x < 0.25 ~ 'a First level',
      x < 0.66 ~ 'b Second level',
      TRUE ~ 'c Last level'
    ),
    groups = str_ord_factor(groups)
  )
    
fct_count(calc_factor$groups)
```

